Ages of Planets and Stars Observed by Kepler

### 2024-04-09 paper/ms.tex
update ms

\begin{figure}[!t]
	\begin{center}
		\leavevmode
		\includegraphics[width=0.49\textwidth]{li_vs_teff_eagles_showdispersion.pdf}
	\end{center}
	\vspace{-0.6cm}
	\caption{
    {\bf Equivalent widths (EW) of the \ion{Li}{1} 6708\,\AA\ doublet
    from Keck/HIRES spectra collected between 2009 and 2024.}
    Lines represent mean models derived from Gaia-ESO spectra of stars
    in clusters of various ages; the intrinsic dispersion 
    evolves in time, and is modelled as the $\pm$1$\sigma$
    bands~\citep{Jeffries_2023}.
    As for rotation-based ages, the intrisic dispersion sets the
    empirical precision limit for the technique.
		\label{fig:li_vs_teff}
	}
\end{figure}

%TODO: does JUMP mirror all HIRES data from KOA?
We queried a combination of the Keck Observatory Archive and the
internal CPS archive to collect all HIRES spectra that were available
for the \nplhoststarwgyroagewithgrazingandhighruwe\ stars
described in Section~\ref{subsec:tally} for which it might be possible
to derive rotation-based ages.

%
We required the spectra to not have the iodine cell in...
%TODO ?!?!?!
%

This exercise yielded at least one spectrum for \nlithiumstars\ and 
\nlithiumplanets\ planets.
These observations were acquired between 6 September 2009 and 18
September 2022.

Out of the \nplhoststarwgyroagewithgrazingandhighruwe\ stars for w

To measure the lithium equivalent widths for the \nkoiswithhires\ KOIs
with HIRES spectra, we use the procedure described by
\citet{Bouma_2021}.
The stars of interest for this work are FGK stars, and so the
continuum in the vicinity of the
\ion{Li}{1} 6708\,\AA\ doublet is well-defined.
We therefore numerically integrate a single best-fit Gaussian over a
local window, and estimate the uncertainties on the line width through
a Monte Carlo procedure that bootstraps against the local scatter in the
continuum.
Generally speaking, this approach does not correct for the neighboring
\ion{Fe}{1} 6707.44\,\AA\ blend.

To evaluate the accuracy and precision of this approach, after
applying an initial iteration of this method on the \nkoiswithhires\
KOIs with HIRES spectra, we compared our lithium equivalent widths
with those reported by \citet{Berger_2018}.
For the \nbergeroverlap\ stars in both samples, 
we found {\bf X, Y, Z.}

To calculate lithium ages, we then use EAGLES \citep{Jeffries_2023}.
{\bf TODO: explain why}
%TODO: explain why.

@@

\subsection{Young Planets}

@@

\startlongtable
\begin{deluxetable}{lllrrrrr}
	%\tabletypesize{\footnotesize}
	\tabletypesize{\scriptsize}
	\tablecaption{Young planets and planet candidates.
		%The non-truncated machine readable versions are accessible both
		%through the online journal, and at
		%\url{https://zenodo.org/record/8327508}.
		\label{tab:youngplanets}}
	%\toprule
	%\midrule
	%\endhead
	\startdata
	KOI & Kepler & $t_{\rm gyro}$ & $R_{\rm p}$ & $P_{\rm orb}$ & $f_{\rm RUWE}$ & $f_{\rm grazing}$ & Spec? \\
  -- &   -- & Myr &    Earths &    days &      bool &   bool  & bool \\
	\hline
	\input{table_subgyr_confirmed} %\\
	%\hline
	\input{table_subgyr_candidate}
	\enddata
	\tablecomments{This table includes X, Y, Z... The machine-readable version,
		available online, includes additional columns for .... 
	}
\end{deluxetable}
%\endlongtable

>>>>

\begin{figure}[!t]
	\begin{center}
		\leavevmode
		\includegraphics[width=0.49\textwidth]{li_vs_teff_koi_X_S19S21dquality_eagles_showdispersion.pdf}
	\end{center}
	\vspace{-0.6cm}
	\caption{
    {\bf Equivalent widths (EW) of the \ion{Li}{1} 6708\,\AA\ doublet
    from Keck/HIRES spectra collected between 2009 and 2024.}
    Lines represent mean models derived from Gaia-ESO spectra of stars
    in clusters of various ages; the intrinsic dispersion 
    evolves in time, and is modelled as the $\pm$1$\sigma$
    bands~\citep{Jeffries_2023}.
    As for rotation-based ages, the intrisic dispersion sets the
    empirical precision limit for the technique.
		\label{fig:li_vs_teff}
	}
\end{figure}

%TODO: does JUMP mirror all HIRES data from KOA?
We queried a combination of the Keck Observatory Archive and the
internal CPS archive to collect all HIRES spectra that were available
for the \nplhoststarwgyroagewithgrazingandhighruwe\ stars
described in Section~\ref{subsec:tally} for which it might be possible
to derive rotation-based ages.
We imposed no constraint on whether the iodine cell was present, since
there are no relevant lines in the region of the 
\ion{Li}{1} 6708\,\AA\ doublet.
This exercise yielded at least one spectrum for \nlithiumstars\ and 
\nlithiumplanets\ planets.
These observations were acquired between 6 September 2009 and 16
May 2024, primarily by the California Kepler Survey (CKS,
CITE).
In preparation for this study, we additionally collected {\bf XX}
spectra for ``confirmed'' planets with rotation-based ages below
1\,Gyr.

To measure the lithium equivalent widths, we followed the procedure
described by \citet{Bouma_2021}.
The stars of interest for this work are FGK stars, and so the
continuum in the vicinity of the
\ion{Li}{1} 6708\,\AA\ doublet is well-defined.
We therefore numerically integrate a single best-fit Gaussian over a
local window, and estimate the uncertainties on the line width through
a Monte Carlo procedure that bootstraps against the local scatter in the
continuum.
Generally speaking, this approach does not correct for the neighboring
\ion{Fe}{1} 6707.44\,\AA\ blend.

To evaluate the accuracy and precision of this approach, after
applying an initial iteration of this method on the 
\nplhoststarwgyroagewithgrazingandhighruwe\ aforementioned stars,
we compared our lithium equivalent widths
with those reported by \citet{Berger_2018}.
For the \nbergeroverlap\ stars in both samples, we found {\bf X, Y,
Z.}

To calculate lithium ages, we use a combination of \texttt{EAGLES}
\citep{Jeffries_2023} and \texttt{BAFFLES}
\citep{Stanford-Moore_2020}.
The former tool was calibrated on stars observed by the Gaia-ESO
spectroscopic survey in 52 open clusters, with ages spanning 2 to
6{,}000\,Myr.  
Example \texttt{EAGLES} isochrones, as well as the modelled intrinsic dispersion
about those isochrones, are shown in Figure~\ref{fig:li_vs_teff}.
While our preference would have been to exclusively use
\texttt{EAGLES} due to its homogeneous analysis approach ...

@@

\subsection{The Sub-Gigayear Sample}

Given that the history of young exoplanet discovery includes a
sociological tendency to report detailed descriptions of specific
young exoplanets, we find it necessary to briefly discuss the
properties of at least the most notable members of the set of
confirmed planets that have well-measured ages.
For this exercise, we follow Table~\ref{tab:youngplanets}.

\paragraph{The Cep-Her Planets}
Kepler\,1627Ab, Kepler\,1643b, Kepler\,1974b and Kepler\,1975\,Ab are
known members of the $\approx$40\,Myr old Cep-Her association
\citep{Bouma_2022a,Bouma_2022b}.
The age of various sub-components of this association are
best measured through relative isochronal placement of the
pre-main-sequence \citep[see][]{Bouma_2022b}, and gyrochronology of
stars in these clusters show them to be $<$100\,Myr old.
The planets themselves are all on close-in orbits (5-25\,days), with
sizes from 2-4\,$R_\oplus$.
For Kepler\,1627A and Kepler\,1974, the rotation and lithium-based
ages for these stars agree within 1$\sigma$ with the cluster-based
ages.
{\bf TODO discuss Kepler-1643 old tension, ditto KOI7913...}

\paragraph{Theia~520 Planets}

Kepler\,968 and Kepler\,52 both have rotation-based ages near
350\,Myr.
They are also members of Theia~520 (CITE).
Independent work by J.~Curtis et al.~is indicating an age of XXX\,Myr for this
cluster.
This makes them the youngest multiplanet systems that have currently
been identified from the Kepler primary mission.

\paragraph{NGC~6811 Planets}
Meibom (CITE CITE) ... do we get those ones?

\paragraph{Systems with Rotation and Lithium in Agreement}
Kepler\,1529b, a 5.3\,day 2.0\,$R_\oplus$ mini-Neptune is
$\approx$100\,Myr old.

Kepler\,1928b, a 19.6\,day 1.9\,$R_\oplus$ mini-Neptune is
$\approx$140-200\,Myr old.

Kepler\,1930b, a 13.0\,day 1.5\,$R_\oplus$ super-Earth on a grazing
orbit is $\approx$120-180\,Myr old.

Kepler\,1313b, doubly unlucky, a 3.8\,day 1.7\,$R_\oplus$ super-Earth
on a grazing orbit is $\approx$185\,Myr old.

Kepler\,1521b, a 47.2\,day 2.3\,$R_\oplus$ super-Earth is 180-210\,Myr
old.

Kepler\,1467b, a 47.1\,day 3.1\,$R_\oplus$ mini-Neptune is
230-240\,Myr old.

Kepler\,1565b, a 1.5\,day 1.3\,$R_\oplus$ super-Earth is 180-230\,Myr
old.

Kepler\,971b, a 9.6\,day 3.9\,$R_\oplus$ Neptune-sized planet is
300-370\,Myr old.

Kepler\,1933b is a 4.9\,day 1.0\,$R_\oplus$ Earth-sized planet.  Its
rotation-based age of $304^{+96}_{-112}$\,Myr agrees within $1\sigma$
of its lithium age, though at half the relative uncertainty.
This system may be the youngest Earth-sized planet known, comparable to 
systems such as HD~63433d \citep[1.1\,$R_\oplus$, $414\pm23$\,Myr][]{2024AJ....167...54C}.

\paragraph{Other Notables}

Kepler~221 is a 633$^{+120}_{-112}$\,Myr four-planet system.

Kepler~51 (CITE Masuda2014) has previously been noted for its bizarre architure.
The rotation-based age we find here is higher/lower than previously noted.

Kepler~411 is also odd, due to the 58\,day-outer planet.

\paragraph{Sub-}

Kepler\,957b

@@

\startlongtable
\begin{deluxetable*}{llllrrrrr}
	%\tabletypesize{\footnotesize}
	\tabletypesize{\scriptsize}
	\tablecaption{Young planets and planet candidates.
		%The non-truncated machine readable versions are accessible both
		%through the online journal, and at
		%\url{https://zenodo.org/record/8327508}.
		\label{tab:youngplanets}}
	%\toprule
	%\midrule
	%\endhead
	\startdata
  KOI & Kepler & $t_{\rm gyro}$ & $t_{\rm Li} $ & $R_{\rm p}$ & $P_{\rm orb}$ & $f_{\rm RUWE}$ & $f_{\rm grazing}$ & Spec? \\
  -- &   -- & Myr & Myr &    Earths &    days &      bool &   bool  & bool \\
	\hline
	\input{table_subgyr_confirmed} %\\
	%\hline
	\input{table_subgyr_candidate}
	\enddata
	\tablecomments{This table includes X, Y, Z... The machine-readable version,
		available online, includes additional columns for .... 
	}
\end{deluxetable*}
%\endlongtable
